---
title: "Data Wrangling Using Dplyr"
author: "TA Crystal"
date: "9/8/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading R packages
* dplyr: dataframe manipulation   
* ggplot2: visualization    
```{r message=FALSE, warning=FALSE}
#install packages only if you have not already done so 
list.of.packages <- c( "dplyr", "tidyverse")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
```


```{r message=FALSE, warning=FALSE}
if(length(new.packages)) install.packages(new.packages)

#library packages
for (pkg in c("dplyr", "tidyverse")) {
  library(pkg, character.only = TRUE)
  }
```


```{r}
load("surgery_data.RData")
```


# I. Mutate Function
## Example: Change the label for a categorical variable
```{r}
anyNA(surgery_data$gender) #check whether there are NA values

table(is.na(surgery_data$gender)) #gives the count of NA values: 3

class(surgery_data$gender) #check the data type 

table(surgery_data$gender) #check how many non-NA levels are there in the gender variable

#overwirte gender variable
surgery_data <- surgery_data%>% 
  mutate(gender = if_else(gender == "F", "Female", 
                  if_else(gender == "M", "Male", "Unknown")))
           
```

## Example: Group patients whose race, gender are NA into a seperate group
There are 480 patients who have NA values for race. We don't want to exclude these sample from our data, let's treat them as a seperate group called "Unknown"
```{r}
anyNA(surgery_data$race) #check whether there are NA values
table(is.na(surgery_data$race)) #gives the count of NA values: 3
table(surgery_data$race)

surgery_data <- surgery_data%>%
  mutate(race = if_else(is.na(race), "Unknown", race))%>%
  mutate(gender = if_else(is.na(gender), "Unknown", gender))

table(surgery_data$race)
table(surgery_data$gender)


anyNA(surgery_data$race)
anyNA(surgery_data$gender)
```



## Example: Create age groups from numeric a age variable
```{r}
anyNA(surgery_data$age) #check whether there are NA values

table(is.na(surgery_data$age)) #gives the count of NA values: 3

class(surgery_data$age) #check the data type 

summary(surgery_data$age) #check the range of the variable

hist(surgery_data$age) #check the distribution of the variable, which helps us to seperate into groups

surgery_data <- surgery_data%>%
  mutate(age_group = if_else(age < 20, "less than 20",
                     if_else(age < 40, "20-40 yrs",
                     if_else(age < 60, "40-60 yrs",
                     if_else(age < 80, "60-80 yrs", 
                             "80+")))))
table(surgery_data$age_group)
```


# II. Select Function
### Example: Only keep variables of interest in the dataframe
```{r}
surgery_data_subset <- surgery_data%>%
  select(age, gender, bmi, hour, race)
glimpse(surgery_data_subset)
```


# III. Filter Function
## Example: Identify one African American patients
```{r}
table(surgery_data$race)

surgery_data_AfricanAmerican <- surgery_data%>%
  filter(race == "African American")
```

## Example: Identify patients who's surgery time is longer than 10 hours
```{r}
hist(surgery_data$hour)

surgery_data_10hr<- surgery_data%>%
  filter(hour > 10)

hist(surgery_data_10hr$hour)
```

# IV. Summarize Function
## Example: Identify the average surgery hour for each race group
```{r}
surgery_data%>%
  group_by(race)%>%
  summarize(count = n(),
            hour_mean = mean(hour),
           hour_median= median(hour),
           hour_sd = sd(hour))%>%
  mutate(perc = count/sum(count) * 100)
```

## Example: Further investigate within each race, what's the average surgery hour for different asa statis
```{r}
table <- surgery_data%>%
  mutate(asa_status = if_else(is.na(asa_status), "Unknown", asa_status))%>%
  group_by(race, asa_status)%>%
  summarize(count = n(),
            hour_mean = mean(hour),
            hour_median= median(hour),
            hour_sd = sd(hour))%>%
  filter(count > 5)

table
```


# V. Arrange Function 
Arrange the median surgery hour in race+asa status group in descending order
```{r}
table%>%
  arrange(-hour_median)
```



