---
title: "Data Wrangling Using Dplyr"
author: "TA Crystal"
date: "9/8/2020"
output:
  pdf_document:
    toc: yes
  html_document:
    theme: cerulean
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading R packages
* dplyr: dataframe manipulation   
* ggplot2: visualization    
```{r message=FALSE, warning=FALSE}
#install packages only if you have not already done so 
list.of.packages <- c( "dplyr", "tidyverse")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
```


```{r message=FALSE, warning=FALSE}
if(length(new.packages)) install.packages(new.packages)

#library packages
for (pkg in c("dplyr", "tidyverse")) {
  library(pkg, character.only = TRUE)
  }
```


```{r}
load("surgery_data.RData")
```


# I. Mutate Function
## Example I.i: Change the label for a categorical variable
Before any data wrangling, we will always perform the 3-step procedure for the variable we are interested in:        

1. check the data type (character or integer or others)

2. check whether there are NAs, how many NAs are there in the variable

3. what values are there in the variable. Use `table()` function for categorical variables, use `summary()` function for numerica variables.   

```{r}
#step 1.
class(surgery_data$gender) #check the data type 

#step 2.
anyNA(surgery_data$gender) #check whether there are NA values
table(is.na(surgery_data$gender)) #gives the count of NA values: 3

#step 3.
table(surgery_data$gender) #check how many non-NA levels are there in the gender variable

#overwirte gender variable
surgery_data <- surgery_data%>% 
  mutate(gender = if_else(gender == "F", "Female", 
                  if_else(gender == "M", "Male", "Unknown")))
           
```

## Example I.ii: Group patients whose race, gender are NA into a seperate group
There are 480 patients who have NA values for race. We don't want to exclude these sample from our data, let's treat them as a seperate group called "Unknown"
```{r}
#step 1.
class(surgery_data$race)

#step 2.
anyNA(surgery_data$race) #check whether there are NA values
table(is.na(surgery_data$race)) #gives the count of NA values: 3

#step 3.
table(surgery_data$race)

surgery_data <- surgery_data%>%
  mutate(race = if_else(is.na(race), "Unknown", race))%>%
  mutate(gender = if_else(is.na(gender), "Unknown", gender))

table(surgery_data$race)
table(surgery_data$gender)


anyNA(surgery_data$race)
anyNA(surgery_data$gender)
```



## Example I.iii: Create age groups from a numeric age variable
```{r}
#step 1
class(surgery_data$age) #check the data type 

#step 2
anyNA(surgery_data$age) #check whether there are NA values
table(is.na(surgery_data$age)) #gives the count of NA values: 3

#step 3.
summary(surgery_data$age) #check the range of the variable
hist(surgery_data$age,
     main = "Distribution of Age",
     xlab = "Age") #check the distribution of the variable, which helps us to seperate into groups

surgery_data <- surgery_data%>%
  mutate(age_group = if_else(age < 20, "less than 20",
                     if_else(age < 40, "20-40 yrs",
                     if_else(age < 60, "40-60 yrs",
                     if_else(age < 80, "60-80 yrs", 
                             "80+")))))
table(surgery_data$age_group)
```


# II. Select Function
### Example II.i: Only keep variables of interest in the dataframe
```{r}
surgery_data_subset <- surgery_data%>%
  select(age, gender, bmi, hour, race)
glimpse(surgery_data_subset)
```


# III. Filter Function
## Example III.i: Identify one African American patients
```{r}
table(surgery_data$race)

surgery_data_AfricanAmerican <- surgery_data%>%
  filter(race == "African American")
```

## Example III.ii: Identify patients who's surgery time is longer than 10 hours
```{r}
hist(surgery_data$hour,
     main = "Distribution of Surgery Duration",
     xlab = "Surgery Duration (hour)")

surgery_data_10hr<- surgery_data%>%
  filter(hour > 10)

hist(surgery_data_10hr$hour,
     main = "Distribution of Surgery Duration(Hour > 10)",
     xlab = "Surgery Duration (hour)")
```

# IV. Summarize Function
## Example IV.i: Identify the average surgery hour for each race group
```{r}
surgery_data%>%
  group_by(race)%>%
  summarize(count = n(),
            hour_mean = mean(hour),
           hour_median= median(hour),
           hour_sd = sd(hour))%>%
  mutate(perc = count/sum(count) * 100)
```

## Example IV.ii: Further investigate within each race, what's the average surgery hour for different asa statis
```{r}
table <- surgery_data%>%
  mutate(asa_status = if_else(is.na(asa_status), "Unknown", asa_status))%>%
  group_by(race, asa_status)%>%
  summarize(count = n(),
            hour_mean = mean(hour),
            hour_median= median(hour),
            hour_sd = sd(hour))%>%
  filter(count > 5)

table
```


# V. Arrange Function 
Arrange the median surgery hour in race+asa status group in descending order
```{r}
table%>%
  arrange(-hour_median)
```



