---
title: "01-Getting Started"
author: "TA Crystal Zang"
date: "8/21/2020"
output:
  html_document:
      theme: cerulean
      toc: true
      toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# I. Vectors
Here we are creating an object and storing values in an object. We say score A is a vector that contains 3 elements, and it's a numeric object. The first element of score A is 85, the second element is 89, and the third element is 92.
```{r eval=FALSE, include=TRUE}
scoreA <- c(85, 89, 92)
scoreA
class(scoreA)

scoreB <- c(75, 80, 95)
scoreB
class(scoreB)
```

# II. Numeric Operations
```{r eval=FALSE, include=TRUE}
1 + 1 

2 * 9 

9/2

9^2

log(9)
```

Now we know the basic numeric operations, we can perform these operations on the vectors we previously created. 
```{r eval=FALSE, include=TRUE}
scoreA + scoreB

#calculate the sum of the elements in scoreA
sum(scoreA)

#find the number of elements in the vector scoreA
length(scoreA)

#calculate the mean of scoreA
mean(scoreA)
sum(scoreA)/length(scoreA)

#calculate standard deviation of scoreA
sd(scoreA)

#inter-quatile range of scoreA
IQR(scoreA)

#summary statistics of scoreA
summary(scoreA)
```

## Question 1. What's the difference in the mean score A and mean score B

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
```{r eval=FALSE, include=FALSE}
#answer
mean(scoreA)-mean(scoreB)
```


# III. Dataframe
**fivethirtyeight** R package hosts many datasets. You can find detailed information on these datasets here: https://fivethirtyeight-r.netlify.app/articles/fivethirtyeight.html. 


We will use **bad_drive** data as an example. The raw data behind the story "Dear Mona, Which State Has The Worst Drivers?" https://fivethirtyeight.com/features/which-state-has-the-worst-drivers/

```{r eval=TRUE, include=TRUE}
#you only install package once
#install.packages("fivethirtyeight")

# install.packages('fivethirtyeightdata', repos ='https://fivethirtyeightdata.github.io/drat/',type = 'source')

#you need to library the package everytime you open this R Markdown 
library(fivethirtyeight)


#gives you the top rows of the dataset
head(bad_drivers)

#data documentation including variable explainations
?bad_drivers

#View drug use dataset in a seperate tab
#View(bad_drivers)
```


There are 51 states and 8 variables in the dataset. 
```{r}
#find the number of row and column
dim(bad_drivers) 
```

```{r}
#using base R to plot the histogram
hist(bad_drivers$perc_speeding)

#useing ggplot2 package to plot the histogram
#install.packages("ggplot2")

library(ggplot2)

ggplot(bad_drivers, aes(x = perc_speeding))+
  geom_histogram(binwidth = 4)  #can manually change binwidth
```



