---
title: "Descriptive Statistics Example"
author: "Arvon Clemons; Crystal Zang"
date: "9/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading R packages
* psych: we are using the summary statistics fucntions provided in this package   
* dplyr: dataframe manipulation   
* ggplot2: visualization    
```{r message=FALSE, warning=FALSE}
#install packages only if you have not already done so 
list.of.packages <- c("psych", "ggplot2", "dplyr")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

#library packages
for (pkg in c("psych", "ggplot2", "dplyr")) {
  library(pkg, character.only = TRUE)
  }
```
Note that the `message = FALSE, warning=FALSE` parameter was added to the code chunk to prevent printing warning and messages when loading the packages.

## Data Description

```{r }
load("surgery_data.RData") #save this data in the same working directory of the rmd file, i.e. in the same folder. You can also click on the Rdata to load it into your RStudio.

#save(surgery_data, file="surgery_data.RData")
glimpse(surgery_data) #previw dataframe, provides data type of each variable: numeric (dbl meaning double precision floating point number), character (chr)

# summary statistics
describe(surgery_data)

#Check for NAs in gender variable
anyNA(surgery_data$gender)

#remove observations with missing 'gender' values using "!" operator and is.na() function
gender_comp <- surgery_data[!is.na(surgery_data$gender), ]

#Create vectors of BMI based on gender; omit NAs
female_bmi <- gender_comp$bmi[gender_comp["gender"] == "F"]
male_bmi <- gender_comp$bmi[gender_comp["gender"] == "M"]

# Calculate mean BMI
# since there may be NA values, use na.rm = T to remove any possible NAs when calculating the mean
mean(female_bmi, na.rm = T); mean(male_bmi, na.rm = T)

#Total numbers of obese or non-obese participants by gender
nonobeseMale <- sum(male_bmi <= 30, na.rm = T)
nonobeseFemale <- sum(female_bmi <= 30, na.rm = T)

obeseMale <- sum(male_bmi > 30, na.rm = T)
obeseFemale <- sum(female_bmi > 30, na.rm = T)

# Calculate proportion of participants who are over 30 BMI
propMale <- obeseMale / (obeseMale + nonobeseMale)
propFemale <- obeseFemale / (obeseFemale + nonobeseFemale)
```

## Visualization

You can also embed plots, for example:

### Bar plot for categorical variables
```{r}
#manually omit NA values in the bar plot using "subset" function
ggplot(data = subset(surgery_data, !is.na(gender)), aes(x = gender))+
  geom_bar() +  #can manually change binwidth
  labs(title = "Gender Frequency in Surgery Data", #label axes
       x = "gender") +
  theme_bw() #make the plot looks pretty

```

### Histogram plot for numeric variables

```{r }
summary(surgery_data$hour) #summmary statistics for 'hour' variable
hour_mean <- mean(surgery_data$hour) #mean 
hour_sd <- sd(surgery_data$hour) #standard deviation

#create a new variable "age_z" in the dataframe, which is the z-score of the age variable
surgery_data$hour_z <- (surgery_data$hour - hour_mean)/hour_sd

#NA values are automatically omitted in the histogram
ggplot(data = surgery_data, aes(x = hour))+
  geom_histogram(bins = 10) +  #can manually change the number of bins, now we have 10 bins
  labs(title = "Distributon of Surgery Time",
       x = "Surgery time (hour)") +
  theme_bw() #make the plot looks pretty
```

Note that we don't have NA values in the hour variable. If there are NAs, use "na.rm=T" argument in the mean and sd calculation.
Ex. "hour_mean <- mean(surgery_data$hour, na.rm=T)"



### Side-by-side Plot for A Numeric Variable by Categories
```{r}
ggplot(data = subset(surgery_data, !is.na(gender)), aes(x = hour))+  
  geom_histogram(bins = 15) +  #can manually change the number of bins, now we have 15 bins
  labs(title = "Distributon of Surgery Hour by Gender",
       x = "Surgery time (hour)") +
  facet_wrap(~gender) + #provides side by side plot by gender
  theme_bw() #make the plot looks pretty
```

### Scatter Plot for Two Numeric Variables 
```{r warning=FALSE}
ggplot(surgery_data, aes(x = age, y = hour))+ 
  geom_point(size = 0.5) +#can adjust the size of the point
   theme_bw() 
```

