---
title: "BIOST 2041 Intro to Statistical Methods"
author: "TA Crystal Zang"
date: "8/23/2020"
output:
  pdf_document:
    toc: yes
  html_document:
    theme: cerulean
    toc: yes
    toc_float: yes
subtitle: R Learning Material 1-- Getting Started
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# I. Basics
Here we are creating a variable and storing values in the variable. We say score A is a vector that contains 3 elements, and it's a numeric variable. The first element of score A is 85, the second element is 89, and the third element is 92. We say the data type of scoreA and scoreB are numeric. Next, we create a student variable, where the student names, Alex, Jordan, and Ryan, are stored. The data type of student vector is character.
```{r eval=TRUE}
#numeric
scoreA <- c()
scoreA <- c(85, 89, 92) #assign values to a variable
    
scoreA #print the variable
class(scoreA) #check the data type of the variable

scoreB <- c(75, 80, 95)
scoreB
class(scoreB)

scoreA[1] < scoreB[1]

1 > 2

scoreB > 80
pass <- scoreB > 80
class(pass)

#character
student <- c("Alex", "Jordan", "Ryan")
class(student)
```


# II. Numeric Operations
```{r eval=FALSE, include=TRUE}
1 + 1 

2 * 9 

9/2

9^2

log(9)
```

Now we know the basic numeric operations, we can perform these operations on the vectors we previously created. 

```{r eval=FALSE, include=TRUE}
scoreA + scoreB

#calculate the sum of the elements in scoreA
sum(scoreA)

#find the number of elements in the vector scoreA
length(scoreA)

min(scoreA) #minimum

max(scoreA) #maximum

#calculate the mean of scoreA
mean(scoreA)
sum(scoreA)/length(scoreA)

#calculate the median of scoreA
median(scoreA)

#calculate standard deviation of scoreA
sd(scoreA)

#inter-quatile range of scoreA
IQR(scoreA)

#summary statistics of scoreA
summary(scoreA)
```

## Question 1. 
#### What's the difference in the mean score A and mean score B

<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
```{r eval=FALSE, include=FALSE}
#answer
mean(scoreA)-mean(scoreB)
```

# III. R package Installation
**fivethirtyeight** R package hosts many datasets. You can find detailed information on these datasets here: https://fivethirtyeight-r.netlify.app/articles/fivethirtyeight.html.   

We need to install an R package using the commend **install.packages()**. You only install package once. But everytime you open you R Markdown file, you have to read in the package using the commend **library()**. To find documentation of any R package use **?** with the name of the R package. For example **?fivethirtyeight**.   

Packages you need to install: **fivethirtyeight**(dataset), **ggplot2**(visualization tool).
```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
#install from a published R package
#install.packages("fivethirtyeight")

#install from the repository to get the most recent updates
#install.packages('fivethirtyeightdata', repos ='https://fivethirtyeightdata.github.io/drat/',type = 'source')

library(fivethirtyeight)
data(bad_drivers)
#gives you the top rows of the dataset
head(bad_drivers)

#data documentation including variable explainations
?bad_drivers

#View drug use dataset in a seperate tab
View(bad_drivers)
```

# IV. Dataframe

We will use **bad_drive** dataframe as an example. The raw data behind the story "Dear Mona, Which State Has The Worst Drivers?" https://fivethirtyeight.com/features/which-state-has-the-worst-drivers/     


First, we want to know the dimension of the dataset. There are 51 U.S. states and 8 variables in the dataset. This dataset is in a "tidied" format, meaning that each row is an observation, preferable unique, and each column is variable. In this dataset, have have each row indicating a state.
```{r}
#find the number of row and column
dim(bad_drivers) 
```

We are interested in the percentage of drivers involved in fatal collisions who were speeding  using the variable **perc_speeding** in the dataset. To extract a variable from a dataframe, we use **$**. An extracted variable is in a vector, similar to the scoreA variable that we showed before. 

```{r}
bad_drivers$perc_speeding #print the variable

length(bad_drivers$perc_speeding) #find the length of the variable

bad_drivers[1, ] #extract one row

```

## Question 2.
#### Report the summary statistics (mean, standard deviation, IQR) of the percent speeding. 
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
```{r eval=FALSE, include=FALSE}
summary(bad_drivers$perc_speeding)
sd(bad_drivers$perc_speeding)
```

### Visualization
Here we are visualizing the distribution of the percentage of drivers involved in fatal collisions who were speeding among all 51 states.  

#### 1. Visualization using Base R
```{r}
#using base R to plot the histogram
hist(bad_drivers$perc_speeding, 
     main = "Distribution of the percent speeding",  #title
     xlab = "percent speeding")  #title of x-axis
```

#### 2. Visualization using ggplot2
```{r}
#useing ggplot2 package to plot the histogram
#install.packages("ggplot2")

library(ggplot2)

ggplot(bad_drivers, aes(x = perc_speeding))+
  geom_histogram(binwidth = 4) +  #can manually change binwidth
  labs(title = "Distribution of the percentage of drivers involved in fatal collisions who were speeding",
       x = "percent speeding") +
  theme(title =element_text(size=9)) +#size of the title
  theme_bw() #make the plot looks pretty
```



## Question 3. 
#### Explore a different variable of your interest in the "bad_drivers" dataset. Calculate the summary statistics and plot the distribution of the variable
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>
<br>


